Use and acquisition of causal relations for decision making under uncertainty using imperfect information repeated games

### 2018-06-14 Propuesta/Proposal_English.tex
actualizando

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
%\usepackage[spanish, mexico]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{listings}
\usepackage{breakcites}
\usepackage{dsfont}
\usepackage{hyperref}
\usepackage{amssymb,amsthm,amsmath,latexsym}
\usepackage[margin=1.5cm]{geometry}
\usepackage{natbib}
%\bibliographystyle{stylename}
%\usepackage{fancyhdr}
%\pagestyle{fancy}
\theoremstyle{plain}
\newtheorem{teo}{Theorem}
\newtheorem{prop}[teo]{Proposition}
\newtheorem{defi}[teo]{Definition}
\newtheorem{obs}[teo]{Observation}
\newtheorem{lem}[teo]{Lemma}
\newtheorem{cor}[teo]{Corolary}
\usepackage[pdftex]{color,graphicx}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{calc}
%\newcommand{\sgn}{\mathop{\mathrm{sgn}}}
\title{Use and acquisition of causal relations for decision making under uncertainty using imperfect information repeated games}
\author{Mauricio Gonzalez Soto}
\begin{document}
%\nocite{*}
\maketitle
\tableofcontents
\newpage
\begin{abstract}
We consider decision problems under uncertainty where the options available to a decision maker and the resulting outcome are related through a causal mechanism which is unknown to the decision maker. We study how a decision maker can learn about this causal mechanism through sequential decision making as well as using current causal knowledge inside each round in order to make better choices had she not considered causal knowledge. We propose a decision making procedure in which an agent holds \textit{beliefs} about her environment which are used to make a choice and then are updated using the observed outcome. As proof of concept, we present an implementation of this causal decision making model and apply it to a simple problem. We show that the model achieves a performance similar to the classic Q-learning while it also acquires a causal model of the environment. 
\end{abstract}
\section{Introduction}
\indent A fundamental part of intelligent reasoning is being able to make decisions under uncertain conditions (\cite{danks2014unifying}, \cite{lake2017building}, \cite{pearlwhy}). In some cases, a decision maker who faces an uncertain environment has enough information to make choices by maximizing expected utility, which is the classic formal criteria for making decisions if rational preferences are assumed (\cite{bernardo2000bayesian}, \cite{gilboa2009decision}). On the other hand, if enough information is not available, the decision maker could attempt to \textit{learn} from the environment by interacting with it.

@@

	In the classic works on Causality (\cite{spirtes2000causation}, \cite{pearl2009causality}) the term \textit{causality} is defined in terms of common language and then the mathematical machinery required to model causality is provided. In this way, causality becomes defined as what is modeled by causal models, which are themselves defined as to model causality. This apparent circularity is deeply studied by \cite{woodward2005making}. For this reason, we will give here a formal working definition of causal relations which allows to be used formally.

@@

	Causality is understood as a relation between \textit{events}. Some event \textit{causes} another event to occurr. The formal definition of Causality that will be adopted in this work is the definition provided in \cite{spirtes2000causation}.

@@

	For interpretations of the axioms in diverse contexts we refer the reader to \cite{bernardo2000bayesian}, \cite{binmore2008rational}, \cite{gilboa2009decision}, \cite{wakker2010prospect}.
	\subsubsection{Subjective Probability and Utility}
	\begin{defi}
	Given a preference relation $\succeq$, we define the (subjective) \textbf{probability} of an event $E \in \mathcal{E}$ as the real number $\mu(S)$ associated with the standard event $S$ such that $E \sim S$.
	\end{defi}
	The subjective probability thus defined satisfies all of the Kolmogorov's axioms of probability (\cite{bernardo2000bayesian}). So, we can use all of the mathematical machinery for classic probability measures, including the definition of conditional probability, Bayes' theorem, etc.
	\begin{defi}
	Consequences $c^\ast$ and $c_\ast$ are called respectively best and worst (extreme consequences) if for any other consequence $c \in \mathcal{C}$ then $c^\ast \succeq c \succeq c_\ast$. Decision problems in which we add extreme consequences are called bounded decision problems.
	\end{defi}
	\begin{defi}
	Given a preference relation $\succeq$ we define the canonical \textbf{utility} $u(c)=u(c | c_\ast, c^\ast)$ of a consequence $c \in \mathcal{C}$ relative to the extreme consequences $c_\ast$, $c^\ast$, as the real number $\mu(S)$ associated with any standard event $S$ such that $c \sim \{c^\ast | S, c_\ast | S^c\}$.
	\end{defi}
	\subsubsection{Decision Criteria for Decision Problems}
	\begin{prop}
	For any bounded decision problem with extreme consequences $c_\ast$ and $c^\ast$ it holds:
	\begin{itemize}
	\item i) For all $c \in \mathcal{C}$, $u(c | c_\ast, c^\ast)$ exists and is unique.
	\item ii) The value of $u(c | c_\ast, c^\ast)$ is unafected by the occurrence of any event $G \succ \emptyset$.
	\item iii) $0 = u(c_\ast | c_\ast, c^\ast) \leq u(c | c_\ast, c^\ast) \leq u(c^\ast | c_\ast, c^\ast) = 1$.
	\end{itemize}
	\end{prop}
	\begin{defi}
	For a bounded decision problem and any event $G \succ \emptyset$ and $a = \{ c_j | E_j : j \in J \}$ we define the conditional expected utility of $a$ as
	\[ \bar{u}(a | c_\ast, c^\ast, G) = \sum_{j \in J} u(c_j | c_\ast, c^\ast)P(E_j | G). \]
	If $G = \Omega$ we simply denote $\bar{u}(a | c_\ast, c^\ast, G)$ as $\bar{u}(a | c_\ast, c^\ast)$
	\end{defi}

@@

	In the particular case where the algegra $\Omega$ contains only the set $\Omega$ we say that it is a decision problem without uncertainty and in this case applying the previous Theorem we obtain the decision criteria which consist in maximizing the utility.

@@

	We follow the exposition from \cite{osborne1994course}, \cite{binmore2007playing}, \cite{shoham2008multiagent}

@@

	It is known that human beings conceive their actions on the world as \textit{intervening} in the world (\cite{hagmayer2009decision}). Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem as if choosing an intervention was choosing an arm of a \textit{slot machine}, in which a gambler chooses an arm and gets some reward. From the rewards they estimate probabilities and output an optimal action in the sense of obtaining minimal regret. Their work considers that the causal model is fully known. They mention that the case where the causal model is unknown is left as an open question, and it is precisely what we are proposing to answer. 
	\subsection{Justification}
	Many real-world applications of decision making are solved by \textit{associative} methods which capture only statistical patterns that are found in data. For example, current methods in Reinforcement Learning, and specifically in Deep Reinforcement Learning, although they have good performance in the task that were supposed to solve, they can not explain \textit{why} a specific trajectory was chosen by the algorithm. This is highly relevant in real-world applications such as self-driving cars, where it is very important to understand why an accident happened. Methods based in Deep Neural Networks aren't supposed to explain why a certain action was chosen since those methods are based in \textit{parallel distributed representations}.

@@

\newpage
\section{Methodology}
\subsection{Modelling a Decision Problem under Uncertainty as a Game}
\subsubsection{Knowing only the graph structure}
\subsubsection{General case}
\subsection{Experimental methodology}
\subsection{Working plan}
\subsubsection{Publication plan}
\begin{itemize}
\item Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML 2018.
\end{itemize}

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
%\usepackage[spanish, mexico]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{listings}
\usepackage{breakcites}
\usepackage{dsfont}
\usepackage{hyperref}
\usepackage{amssymb,amsthm,amsmath,latexsym}
\usepackage[margin=1.5cm]{geometry}
\usepackage{natbib}
%\bibliographystyle{stylename}
%\usepackage{fancyhdr}
%\pagestyle{fancy}
\theoremstyle{plain}
\newtheorem{teo}{Theorem}
\newtheorem{prop}[teo]{Proposition}
\newtheorem{defi}[teo]{Definition}
\newtheorem{obs}[teo]{Observation}
\newtheorem{lem}[teo]{Lemma}
\newtheorem{cor}[teo]{Corolary}
\usepackage[pdftex]{color,graphicx}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{calc}
%\newcommand{\sgn}{\mathop{\mathrm{sgn}}}
\title{Use and acquisition of causal relations for decision making under uncertainty using imperfect information repeated games}
\author{Mauricio Gonzalez Soto}
\begin{document}
%\nocite{*}
\maketitle
\tableofcontents
\newpage
\begin{abstract}
We consider decision problems under uncertainty where the options available to a decision maker and the resulting outcome are related through a causal mechanism which is unknown to the decision maker, although he is aware of the causal nature of his environment. We study how a decision maker can learn about this causal mechanism through sequential decision making as well as using current causal knowledge inside each round in order to make better choices had he not considered causal knowledge. We propose a decision making procedure in which an agent holds \textit{beliefs} about her environment which are used to make a choice and then are updated using the observed outcome. As proof of concept, we present an implementation of this causal decision making model and apply it to a simple problem. We show that the model achieves a performance similar to the classic Q-learning while it also acquires a causal model of the environment. 
\end{abstract}
\section{Introduction}
\indent A fundamental part of intelligent reasoning is being able to make decisions under uncertain conditions (\cite{danks2014unifying}, \cite{lake2017building}, \cite{pearlwhy}). In some cases, a decision maker who faces an uncertain environment has enough information to make choices by maximizing expected utility, which is the classic formal criteria for making decisions if rational preferences are assumed (\cite{bernardo2000bayesian}, \cite{gilboa2009decision}). On the other hand, if enough information is not available, the decision maker could attempt to \textit{learn} from the environment by interacting with it.

@@

	In the reference works on Causality (\cite{spirtes2000causation}, \cite{pearl2009causality}) the term \textit{causality} is defined in terms of common language and then the mathematical machinery required to model causality is provided. In this way, causality becomes defined as what is modeled by causal models, which are themselves defined as to model causality. This apparent circularity is deeply studied by \cite{woodward2005making}. For this reason, we will give here a formal working definition of causal relations which allows to be used formally.

	Several definitions of Causality exist in diverse contexts with diverse applications in mind, such as the Topological Causality for Dynamical Systems (\cite{harnack2017topological}), Lamport's Causality (\cite{lamport1978time}), Granger's Causality for Time Series (\cite{granger1969investigating}), Suppes' Causality (\cite{suppes1970probabilistic}) each with its own mathematical framework. A review of several of this theories, including Suppes' and Granger's can be found in \cite{holland1985statistics}.

@@

	Causality is understood as a \textit{stochastic} relation between \textit{events}. Some event \textit{causes} another event to occurr. For example, The formal definition of Causality that will be adopted in this work is the definition provided in \cite{spirtes2000causation}.

@@

	For interpretations and critiques of the axioms we refer the reader to \cite{bernardo2000bayesian}, \cite{binmore2008rational}, \cite{gilboa2009decision}, \cite{wakker2010prospect}, \cite{peterson2017introduction}.
	\subsubsection{Subjective Probability and Utility}
	\begin{defi}
	Given a preference relation $\succeq$, we define the (subjective) \textbf{probability} of an event $E \in \mathcal{E}$ as the real number $\mu(S)$ associated with the standard event $S$ such that $E \sim S$.
	\end{defi}
	The subjective probability thus defined satisfies all of the Kolmogorov's axioms of probability (\cite{bernardo2000bayesian}). So, we can use all of the mathematical machinery for classic probability measures, including the definition of conditional probability, Bayes' theorem, etc.
	\begin{defi}
	Consequences $c^\ast$ and $c_\ast$ are called respectively best and worst (extreme consequences) if for any other consequence $c \in \mathcal{C}$ then $c^\ast \succeq c \succeq c_\ast$. Decision problems in which we add extreme consequences are called bounded decision problems.
	\end{defi}
	\begin{defi}
	Given a preference relation $\succeq$ we define the canonical \textbf{utility} $u(c)=u(c | c_\ast, c^\ast)$ of a consequence $c \in \mathcal{C}$ relative to the extreme consequences $c_\ast$, $c^\ast$, as the real number $\mu(S)$ associated with any standard event $S$ such that $c \sim \{c^\ast | S, c_\ast | S^c\}$.
	\end{defi}
	\subsubsection{Decision Criteria for Decision Problems}
	\begin{prop}
	For any bounded decision problem with extreme consequences $c_\ast$ and $c^\ast$ it holds:
	\begin{itemize}
	\item i) For all $c \in \mathcal{C}$, $u(c | c_\ast, c^\ast)$ exists and is unique.
	\item ii) The value of $u(c | c_\ast, c^\ast)$ is unafected by the occurrence of any event $G \succ \emptyset$.
	\item iii) $0 = u(c_\ast | c_\ast, c^\ast) \leq u(c | c_\ast, c^\ast) \leq u(c^\ast | c_\ast, c^\ast) = 1$.
	\end{itemize}
	\end{prop}
	\begin{defi}
	For a bounded decision problem and any event $G \succ \emptyset$ and $a = \{ c_j | E_j : j \in J \}$ we define the conditional expected utility of $a$ as
	\[ \bar{u}(a | c_\ast, c^\ast, G) = \sum_{j \in J} u(c_j | c_\ast, c^\ast)P(E_j | G). \]
	If $G = \Omega$ we simply denote $\bar{u}(a | c_\ast, c^\ast, G)$ as $\bar{u}(a | c_\ast, c^\ast)$
	\end{defi}

@@

	In the particular case where the algebra $\mathcal{E}$ contains only the set $\Omega$ as its only element we say that it is a decision problem without uncertainty and in this case applying the previous Theorem we obtain the decision criteria which consist in maximizing the utility function.

@@

	We follow the exposition from \cite{osborne1994course}, \cite{binmore2007playing}, \cite{shoham2008multiagent}.

@@

	It is known that human beings conceive their actions on the world as \textit{intervening} in the world (\cite{hagmayer2009decision}). Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem as if choosing an intervention was choosing an arm of a \textit{slot machine}, in which a gambler chooses an arm and gets some reward. From the rewards they estimate probabilities and output an optimal action in the sense of obtaining minimal regret. Their work considers that the causal model is fully known. They mention that the case where the causal model is unknown is left as an open question, and it is precisely what we are proposing to answer. 
	\subsection{Justification}
	Many real-world applications of decision making are solved by \textit{associative} methods which capture only statistical patterns that are found in data. For example, current methods in Reinforcement Learning, and specifically in Deep Reinforcement Learning, although they have good performance in the task that were supposed to solve, they can not explain \textit{why} a specific trajectory was chosen by the algorithm. This is highly relevant in real-world applications such as self-driving cars, where it is very important to understand why an accident happened.

	For example, as told by \cite{bornstein2016artificial}, at the University of Pittsburgh Medical Center a team of researchers tried to use Machine Learning to predict whether pneumonia patients might develop severe complications. For this purpose they trained Neural Networks and Decision Trees using the hospital's own data. Neural Networks outperformed Decision Trees, but only by studying the decisions made by the latter did the doctors find out that the algorithms instructed doctors to send home pneumonia patients who already had asthma, despite the fact that asthma sufferers are known to be extremely vulnerable to complications. The problem relied in the training data, because the hospital policy was to automatically send asthma sufferers with pneumonia to intensive care, and this policy worked so well that asthma sufferers almost never developed severe complications. It was only through the interpretability of the Decision Trees that the doctors didn't send asthma patients with pneumonia home to a certain death.

	Methods based in Deep Neural Networks aren't supposed to explain why a certain output was produced since those methods are based in \textit{parallel distributed representations} and the same goes for any other learning algorithm that uses Deep Neural Networks, like Deep Reinforcement learning; when AlphaGo (\cite{silver2017mastering}) defeated the world-champion, humanity didn't learn anything new about the game of Go, because the algorithm was not designed to explain its moves, it was just curve fitting.

@@

\newpage
\section{Methodology}
\subsection{Modelling a Decision Problem under Uncertainty as a Game}
\subsubsection{Fully knowing the causal graphical model}
\subsubsection{Knowing only the graph structure}
\subsubsection{General case: Model unknown}
\subsection{Validation of the proposed methods}
\subsection{Experimental methodology}
\subsection{Working plan}
\subsubsection{Publication plan}
\begin{itemize}
\item Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML 2018.
\end{itemize}
